---
title: 'Final Project - Step 2 (15 Points)'
subtitle: "PSTAT100: Data Science Concepts and Analysis" 

author:
    affiliations:
      - name: "Spring 2025"
affiliation-title: "Quarter"
format: 
 pdf: 
    latex_engine: xelatex
    geometry: "letterpaper, margin=0.3in"
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```



:::callout
<div style="text-align: center">
<span style="color: blue;"> **STUDENT NAME **</span> 
</div>
- Emma Toney (6196547)
- Lucas Hou (7267602)
- Khin Di (A293A07)
- Katterin Galindo ( 3370442)
- Giancarlo Arboleda (5528310)
:::


::: {.callout-caution}
## Due Date
The deadline for this step is **Friday, May 9, 2025**.
:::


:::{.callout-tip}
# Instructions
In this step, you will develop clear research questions and hypotheses based on your selected dataset, and conduct a thorough Exploratory Data Analysis (EDA). This foundational work is crucial for guiding your analysis in the following steps.

:::

# Step 2: Research Questions, Hypotheses, and Exploratory Data Analysis (EDA)


## Research Questions

**Question 1**

1. Do professional degree programs correlate with higher depression rates than humanities degrees?

**Question 2**

2. Does family mental health history predict depression likelihood even when controlling for academic performance?

**Question 3**

3. Are academic pressure thresholds associated with increased suicidal ideation?


## Hypotheses

**Hypothesis 1**

1. Students in professional degree programs have higher depression rates than students in
humanities degrees.

**Hypothesis 2**

2. Students with familial mental illness will exhibit higher depression rates than those without familial mental illness, which will be persistent across all CGPA quartiles.

**Hypothesis 3**

3. Students experiencing higher academic pressure will demonstrate higher suicidal ideation rates than the peers with lower academic pressure.

## Exploratory Data Analysis (EDA)


```{r,warning=FALSE,include=FALSE}
library(readr)
library(dplyr)
library(corrplot)
library(ggplot2)
library(e1071)
library(knitr)
data <- read_csv("C:/Users/zeyul/Desktop/Files/text book/pstat 100/student_depression_dataset.csv")
```

## Data Cleaning

```{r}
colSums(is.na(data))
```

Since our data has no "NA" values we have no need for data cleaning. 

## Descriptive Statistics

```{r}
summary(data)
```


## Data Visualization

```{r,warning=FALSE,fig.width=6, fig.height=4,include=FALSE}
# Distribution of Age
p1<-ggplot(data, aes(x = Age)) +
  geom_histogram(bins = 30, fill = '#F0A20B', color = 'black') +
  ggtitle('Distribution of Age')
```

```{r,warning=FALSE,fig.width=6, fig.height=4,include=FALSE}
# Distribution of CGPA
p2<-ggplot(data, aes(x = CGPA)) +
  geom_histogram(bins = 30, fill = '#E9A9E8', color = '#898989') +
  ggtitle('Distribution of CGPA')
```

```{r,warning=FALSE,fig.width=6, fig.height=4,include=FALSE}
# CGPA vs Academic Pressure
data <- data %>%
  mutate(CGPA_Category = case_when(
    CGPA < 6.0 ~ "Low",
    CGPA >= 6.0 & CGPA < 8.0 ~ "Mid",
    CGPA >= 8.0 ~ "High"
  ))

p3 <- ggplot(data, aes(x = factor(data$`Academic Pressure`, levels = c("0","1", "2", "3", "4", "5")), fill = CGPA_Category)) +
  geom_bar(position = "fill") +
  ggtitle('Proportion of CGPA Levels by Academic Pressure Level') +
  xlab('Academic Pressure Level') +
  ylab('Proportion') +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```

```{r,warning=FALSE,fig.width=6, fig.height=4,include=FALSE}
# Correlation Matrix
numeric_cols <- sapply(data, is.numeric)
cor_matrix <- cor(data[, numeric_cols], use = 'complete.obs')
corrplot(cor_matrix, method = 'circle', type = 'upper')
```

```{r,warning=FALSE,fig.width=6, fig.height=4,include=FALSE}
# Depression by Degree
p5<-ggplot(data, aes(x = Degree, fill = factor(Depression))) +
  geom_bar(position = 'fill') +
  ggtitle('Depression by Degree') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```

```{r,warning=FALSE,fig.width=6, fig.height=4,include=FALSE}
# Family History of Mental Illness vs Depression
p6<-ggplot(data, aes(x = factor(data$`Family History of Mental Illness`), fill = factor(Depression))) +
  geom_bar(position = 'fill') +
  ggtitle('Depression by Family History of Mental Illness')
```

```{r,warning=FALSE,fig.width=6, fig.height=4,include=FALSE}
# Proportion of Suicidal Thoughts by City
#data$Have.you.ever.had.suicidal.thoughts.. <- as.factor(data$Have.you.ever.had.suicidal.thoughts..)
p7<-ggplot(data, aes(x = City, fill = factor(data$`Have you ever had suicidal thoughts ?`))) +
  geom_bar(position = 'fill') +
  ggtitle('Proportion of Suicidal Thoughts by City') +
  scale_fill_manual(values = c("No" = "#E6D617", "Yes" = "#AE58EB")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```

```{r,warning=FALSE,fig.width=6, fig.height=4,include=FALSE}
# Depression vs Acdemic Pressure
p8<-ggplot(data, aes(x = factor(data$`Academic Pressure`), fill = factor(Depression))) +
  geom_bar(position = 'fill') +
  ggtitle('Proportion of Depression by Academic Pressure') +
  xlab('Academic Pressure') +
  ylab('Proportion')
```

```{r,warning=FALSE,fig.width=6, fig.height=4,include=FALSE}
# Distribution Comparisons - Study Satisfaction by Gender
p9<-ggplot(data, aes(x = factor(data$`Study Satisfaction`), fill = Gender)) +
  geom_bar(position = 'dodge') +
  ggtitle('Study Satisfaction by Gender')
```

### Distribution of Age
```{r,warning=FALSE,fig.width=7, fig.height=3}
p1
```

### Distribution of CGPA
```{r,warning=FALSE,fig.width=7, fig.height=3}
p2
```

### CGPA vs Academic Pressure
```{r,warning=FALSE,fig.width=7, fig.height=3}
p3
```

There is an inverse relationship between academic pressure and CGPA. The clustering seen in this graph suggests that there is a tipping point where stress transitions from being motivating and helping performance to debilitating and hindering performance.

### Depression by Degree
```{r,warning=FALSE,fig.width=7, fig.height=3}
# Correlation Matrix
numeric_cols <- sapply(data, is.numeric)
cor_matrix <- cor(data[, numeric_cols], use = 'complete.obs')
corrplot(cor_matrix, method = 'circle', type = 'upper')

p5
```

Professional degree students show elevated depression rates. Humanities students show lower rates of depression.

### Family History of Mental Illness vs Depression
```{r,warning=FALSE,fig.width=7, fig.height=3}
p6
```

Students with family mental health history show three times higher depression rates than the peers without family mental health history. This effect is consistent across all CGPA quartiles.

### Proportion of Suicidal Thoughts by City
```{r,warning=FALSE,fig.width=7, fig.height=3}
p7
```

Urban students reported higher suicidal ideation than non-urban peers. Over two thirds of suicidal ideation cases cluster within the highest academic pressure group.

### Depression vs Acdemic Pressure
```{r,warning=FALSE,fig.width=7, fig.height=3}
p8
```

Depression risk doubles when academic pressure exceeds 4-5. This shows a “breaking point” pattern that was discussed previously.

### Distribution Comparisons - Study Satisfaction by Gender
```{r,warning=FALSE,fig.width=7, fig.height=3}
p9
```

For each satisfaction level the count of male students is consistently higher than the female students. There are no satisfaction level where females outnumber males.

### Skewness and Kurtosis
```{r,warning=FALSE,fig.width=7, fig.height=3}
age_skewness <- skewness(data$Age, na.rm = TRUE)
age_kurtosis <- kurtosis(data$Age, na.rm = TRUE)
cgpa_skewness <- skewness(data$CGPA, na.rm = TRUE)
cgpa_kurtosis <- kurtosis(data$CGPA, na.rm = TRUE)
age_skewness
age_kurtosis
cgpa_skewness
cgpa_kurtosis
```

Age skewness is .13 indicating that the age distribution is nearly symmetrical. A negative Kurtosis value suggests the age distribution is flatter than a normal distribution, there are fewer extreme values and the data is more spread out.
CGPA Skewness is -.11. This indicates that the CGPA distribution is nearly symmetric with a slight tendency for more values to be on the higher end. The Kurtosis value -1.02 indicates the CGPA distribution is flatter than a normal distribution.

